Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-06-04 Manuscript/manuscript.Rmd
edits

@@

<!--![Caption for the picture.](/path/to/image.png)-->

@@

<!--![Caption for the picture.](/path/to/image.png)-->

@@

<!--We then went on the evaluate the effect of class imbalance on model performance.--> In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive to models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7).

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the granularity [?? Really confused by the constant use of granularity. Why is it needed ??] by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive for models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7). [??? Need to know why PRCs give different results.]

@@

<!--![Caption for the picture.](/path/to/image.png)-->

@@

<!--![Caption for the picture.](/path/to/image.png)-->

@@

<!--![Caption for the picture.](/path/to/image.png)-->

@@

<!--![Caption for the picture.](/path/to/image.png)-->

@@

<!--![Caption for the picture.](/path/to/image.png)-->

>>>>

**Table 2. Inter-Resolution ICCs.** For GM12878, two-way intraclass correlation coefficients between resolutions for the following quantities: Number of genomic bins, number of TADs called, number of unique TAD boundaries, and imbalance ratio. The ICC was computed across 21 of the autosomal chromosomes. P-values indicate the level of significance of the correlation between data resolution and each quantity.[??? What is ICC? !!!editted]

| Quantity                 | Inter-Resolution | 95% CI          | P-Value  |
|--------------------------|------------------|-----------------|----------|
| Number of Genomic Bins   | 0.464            | [0.244, 0.690]  | < 0.0001 |
| Number of TADs Called    | 0.627            | [0.426, 0.802]  | < 0.0001 |
| Number of TAD Boundaries | 0.611            | [0.407, 0.791]  | < 0.0001 |
| Imbalanced Ratio         | 0.114            | [-0.058, 0.372] | 0.109    |

@@

![**Predictor types.** Three separate types of predictors were considered when quantifying the association between TAD boundaries and genomic annotations. Each predictor type was used to build the feature space of specific models in the ensemble framework. (A) _Overlap Counts_ (OC) consider the total number of genomic features that overlapped with each genomic bin. (B) _Overlap Percent_ (OP) was calculated by dividing the sum of all feature widths within each specific bin by the total bin width. (C) _Distance_ ($log_{2}$ transformed) was calculated by measuring the distance (in bases) from the center of each genomic bin to the center of the nearest region defining a genomic annotation. ](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure2_predictor_types.png)

@@

![**Figure 3. Ensemble Framework.** A diagram of the cell line specific ensemble framework detailing the combination of inputs used in the model building pipeline. The minoirty and majority classes were represented by the genomic bins that overlapped with flanked TAD boundaries or did not, respectively. Functional genomic annotations obtained from the UCSC Genome Browser Database composed the feature space ($\textbf{X} = \{X_1, X_2, ..., X_p\}$). A feature engineering step was included to isolate the predictor-type specific spatial relationship between annotations and genomic bins in increasing granularity, denoted by the red gradient. The data was split into a 7:3 training to testing set ratio. Data re-sampling was then applied to the training set. Lastly, a random forest classification algorithm was implemeted and validated on the testing data.](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure3_ensemble_framework.png)

@@

<!--We then went on the evaluate the effect of class imbalance on model performance.--> In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 4A). We gradually increased the granularity [?? Really confused by the constant use of granularity. Why is it needed ??] by assessing models built on OP predictor types, while still not performing any data resampling. Here, the results were similar to that of OC types (Figure 4B). However, when including distance-type predictors, models were found to exhibit a marked improvement in MCC compared to the other two types (Figure 4C). Results were found to be consistently poor across resolutions and predictor types, when implemented on the K562 cell line (Supplementary Figure 3). Furthermore, we found that conventional metrics of performance (i.e. accuracy, sensitivity and specificity, and AUC) were less sensitive to models built on imbalanced data, regardless of cell line, resolution, or predictor-type (Supplementary Tables 6-7).

@@

![**Figure 4. SMOTE re-sampled models built on distance-type predictors yielded optimal MCCs across resolutions for GM12878.** Performances of random forest classification models built on each of the 3 predictor types: (A) OC, (B) OP, and (C) Distance. Within each predictor type, and across resolution, training data was re-sampled using different techniques including: None (red), ROS (green), RUS (blue), and SMOTE (purple). Performances were aggregated using means across each chromosome, with error bars representing 1 standard deviation away from the mean.](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure4_gm12878_model_performances.png)

@@

![**Figure 5. The proportional importance of optimally built predictive models was outweighed by transcription factor binding sites across resolutions for GM12878.** Stacked barplots illustrate the proportion of total predictive importance in models using optimal data level characteristics determined by the ensemble framework (SMOTE re-sampling; with distance-type predictors), for each resolution on GM12878. Proportions are grouped by the class of cell line specific annotations in the feature space including chromatin state segmentations (CSS), histone modifications (HM), and transcription factor binding sites (TFBS).](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure5_proportion_of_importances_gm12878.png)

@@

![**Figure 6. Optimally built TFBS-specific models outperformed other annotation-specific models across resolution for GM12878** Performances were compared for random forest models built using specific groups of annotations as the feature space including: histone modifications (HM; red), chromatin state segmentations (CSS; green), transcription factor binding sites (TFBS; blue), and a feature space containing all 3 types combined (combined; purple). Each bar represents a model ran through the ensemble framework, with re-sampling technique and predictor-type characteristics fixed to SMOTE and distance-types. The error bars represent $\pm$ 1 standard deviation from the mean MCC across each chromosome.](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure6_comparing_sets_of_annotations_gm12878.png)

@@

![**Figure 7. Top 20 most predictive transcription factor binding sites sorted by 10 kb for GM12878.** A heatmap illustrating the importance of the key cell line specific transcription factors for optimally built random forest models across resolution. Importances were sorted according to the 10 kb resolution and rescaled to values between 0 and 100. Transcription factors with higher values of predictive importance are denoted in dark red.](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure7_variable_importance_heatmaps_tfbs_gm12878.png)

@@

![**Figure 8. Top predictive transcription factor binding sites were significantly closer to bins that overlapped with 10 kb flanked TAD boundaries on GM12878** Boxplots illustrating the distribution of distances (on $log_{2}$ scale) around 10 kb regions on the linear genome that either overlapped with a flanked TAD boundary (Yes) or did not (No), for the most influential cell line specific TFBS including SMC3, CTCF, RAD21, ZNF143, and YY1sc281. P-values obtained from Wilcoxon Rank Sum tests are provided assessing the significance [$NS$ = p > 0.05; $*$ = 0.01 < p $\le$ 0.05; $**$ = 0.001 < p $\le$ 0.01; $***$ = 0.0001 < p $\le$ 0.001; $****$ = p $\le$ 0.00001].](C:/Users/stili/Documents/TAD_predictive_modeling/Manuscript/Figures/figure8_distance_boxplots_key_tfbs_gm12878_10kb.png)
